MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-06-03 main/MSstatsPTM_Main.tex
Intro and Data

%% for comments
\newcommand{\ignore}[1]{}
\def\todo#1{{\color{red}[#1]}}
\def\change#1{{\color{cornflowerblue}#1}}
\newenvironment{todolong}{\color{red}[TODO:}{]}
\def\note#1{{\color{OliveGreen}[NOTE: #1]}}
\def\added#1{{\color{blue}[ADDED: #1]}}
\def\devon#1{{\color{green}[4Devon: #1]}}

@@

\title{MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments}

@@

Liquid chromatography coupled with mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) between samples from various conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of representative peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Localizing the modification sites, or characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.

@@

The technological difficulties in PTM identification and quantification challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. In particular, methods such as $t$-test, Analysis of Variance, or Limma take as input the intensity ratios of modified and unmodified peptide features, and compare the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such methods do not fully account for all the sources of uncertainty and variation in the experiment. As the result, they are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

This manuscript proposes a general statistical analysis framework that detects relative changes in post-translational modifications. The framework requires an experimental design, which for each sample quantifies both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, acquisition methods, and experimental designs, showing the applicability of the framework to a variety of situations. The evaluations demonstrated that by appropriately leveraging the information from the unmodified peptides, the proposed approach improves the accuracy of the estimates of PTM fold changes and results in a better calibrated false positive rate of detecting differentially abundant PTMs. In particular, accounting for the confounding from protein abundance allows us to characterize the true effect of the modification, avoiding the need for more manual and time intensive methods.

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as $MSstats$ and $MSstatsTMT$ \cite{Choi:2014,Huang:2020}.

\section{Experimental Procedures}

The datasets in this manuscript both benchmarked the proposed approach in situations with known ground truth and represented a variety of experimental designs and acquisition methods. Two computer simulations varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. The spike-in experiment took the real world characteristics a step further and allowed us to compare the methods in a real-world experimental settings with known changes in modified spike-in peptides. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies. Table \ref{tab:dataDescription} summarizes the experiments.

\subsection*{Data Availability}

Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
}
\label{tab:dataDescription}
\end{table}

The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).

\subsection*{Dataset 1 : Computer simulation 1 - Label-free}
\label{sec:comp_sim_procedure1}

The simulation represented an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM; each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no difference between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All differential PTMs were generated with an expected log fold change of 0.75 between conditions.

Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$, motivated by the biological experiments in this manuscript.
Further details are in Supplementary Sec. \ref{supp-sec:dataset1}.

\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low features}
\label{sec:comp_sim_procedure2}

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. Specifically, PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. Further details are in Supplementary Sec. \ref{supp-sec:dataset2}.

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}

In the custom spike-in experiment 8 replicates were allocated to 4 conditions, resulting in 2 replicates per condition.

\medskip \noindent {\bf Experimental Design:} 50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark. An overview of the experimental design is shown in Figure \ref{fig:benchmark-design}. Four mixture conditions were created with differing amounts of human lysate, background lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were used to estimate changes in global protein abundance. Background {\it E. coli} lysate was used to normalize total protein levels prior to enrichment or global protein profiling. The background lysate is expected to change in the same way as the corresponding protein. Thus it is treated as control and is expected to show no fold change in all comparisons after adjusting for changes in the unmodified protein abundance. The true log fold changes of spike-in modified peptides in each comparison can be seen in Figure \ref{fig:benchmark-design}(b). Two sets of data were acquired for each mixture in label-free LC-MS/MS mode: KGG enriched + LC-MS/MS, and LC-MS/MS only.

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run to measure unmodified peptides. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect known changes in abundance between pairs of conditions. The four mixtures were treated as individual conditions and labeled mix1, mix2, mix3, and mix4. The pairwise comparisons were mix1-mix2, mix1-mix3, mix1-mix4, mix2-mix3, mix2-mix4, and mix3-mix4.

%\begin{table}[h!]
%\centering
%\begin{tabular}{| c | c |}
%\hline
% Comparison & Expected log fold change\\ %[0.5ex]
% \hline\hline
% mix1-mix2 & -1\\
% \hline
% mix1-mix3 & 1\\
%\hline
% mix1-mix4 & 0\\
%\hline
% mix2-mix3 & -2\\
%\hline
% mix2-mix4 & 1\\
%\hline
% mix3-mix4 & -1\\
%\hline
%\end{tabular}
%\caption{The expected log fold change of %modified spike-in peptides in Dataset 3.}
%\label{table:spikein_fold_change}
%\end{table}

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}
In this experiment 11 biological replicates were allocated to 1 TMT mixture in a group comparison unbalanced design. Further details are in Supplementary Sec. \ref{supp-sec:ipah}.

\medskip \noindent {\bf Experimental Design:} Luchetti et al. \cite{LUCHETTI2021} quantified the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The proteome was quantified using isobaric multiplexing via tandem mass tagging (TMT) in combination with one LC-MS/MS run.

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run, with the same design, to measure unmodified peptides. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Pairwise Comparisons:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other, resulting in 15 pairwise comparisons. Since the dataset was a biological investigation, the true positive modifications were unknown.

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}
\todo{For each experimental dataset, could you say how many proteins were quantified, and the average number of {SM per protein (modified/unmodified)? I was looking everywhere and could not find the info} In this experiment 8 biological replicates were allocated to 2 TMT mixtures in an unbalanced time course design. More details are in Supplementary Sec. \ref{supp-sec:shigella}.

\medskip \noindent {\bf Experimental Design:} Maculins et al. \cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S. flexneri}). The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and uninfected with {\it S. flexneri}. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).

The proteome was quantified in 11-plex isobaric multiplexing via tandem mass tagging (TMT) in combination with LC-MS/MS. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT.

\medskip \noindent {\bf Global Profiling Run:} This experiment included a separate global profiling run with the same design to measure unmodified peptides. There was a 90\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

@@

\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}
\label{sec:exp_proc_dataset6}

In this experiment 8 biological replicates were split between 4 conditions in a balanced group comparison design. Further details in Suppler Sec \ref{supp-sec:usp30}.

>>>>

%% for comments
\newcommand{\ignore}[1]{}
\def\todo#1{{\color{red}[#1]}}
\def\change#1{{\color{cornflowerblue}#1}}
\newenvironment{todolong}{\color{red}[TODO:}{]}
\def\ov#1{{\color{red}#1}}
\def\note#1{{\color{OliveGreen}[NOTE: #1]}}
\def\added#1{{\color{blue}[ADDED: #1]}}
\def\devon#1{{\color{green}[4Devon: #1]}}

@@

\title{MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics}

@@

Liquid chromatography coupled with bottom up mass spectrometry (LC-MS/MS)-based proteomics is increasingly used to detect changes in post-translational modifications (PTMs) in samples conditions. Analysis of data from such experiments faces numerous statistical challenges. These include the low abundance of modified proteoforms, the small number of observed peptides that span modification sites, and confounding between changes in the abundance of PTM and the overall changes in the protein abundance. Therefore, statistical approaches for detecting differential PTM abundance must integrate all the available information pertaining to a PTM site, and consider all the relevant sources of confounding and variation. In this manuscript we propose such a statistical framework, which is versatile, accurate, and leads to reproducible results. The framework requires an experimental design, which quantifies, for each sample, both peptides with post-translational modifications and peptides from the same proteins with no modification sites. The proposed framework supports both label-free and tandem mass tag (TMT)-based LC-MS/MS acquisitions. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, by fitting separate linear mixed effects models appropriate for the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources. Evaluations on computer simulations, a spike-in experiment with known ground truth, and three biological experiments \ov{with different organisms, modification types and data acquisition types} demonstrate the improved fold change estimation and detection of differential PTM abundance, as compared to currently used approaches. The proposed framework is implemented in the free and open-source R/Bioconductor package $MSstatsPTM$.

@@

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}.

@@

The technological difficulties in PTM identification and quantification increase the uncertainty and the variation in the data, and challenge the downstream statistical analyses. Frequently data from these experiments are analyzed using statistical methods that were not originally designed for this task. Researchers use methods such as $t$-test \todo{ref}, Analysis of Variance\todo{ref}, or Limma\todo{ref}, by taking as input the intensity ratios of modified and unmodified peptide features, and comparing the mean abundance of different PTM sites~\cite{Ritchie_15a}. Such approaches do not fully account for all the sources of uncertainty. As the result, these approaches are either not directly applicable to experiments with non-trivial designs (such as experiments with multiple conditions, paired and time course designs, and experiments with labeling), or require the analysts to exercise non-trivial statistical expertise.

This manuscript proposes a versatile statistical analysis framework that accurately detects relative changes in post-translational modifications. The framework requires an experimental design, which quantifies, for each sample, both the peptides with post-translational modifications, and peptides from the same proteins with no modification sites. The framework supports data-dependent acquisitions (DDA) that are label-free or tandem mass tag (TMT)-based. The statistical methodology separately summarizes the abundances of peptides with and without the modification sites, and fits separate linear mixed effects models that reflect the biological and technological aspects of the experimental design. Next, model-based inferences regarding the PTM and the protein-level abundances are combined to account for the confounding between these two sources.

We evaluated the proposed framework on two datasets from computer simulations, one benchmark controlled mixture, and three biological investigations. The datasets illustrate a diverse set of organisms, modification types, acquisition methods and experimental designs, showing the applicability of the framework to a variety of situations. By appropriately leveraging the information from the unmodified peptides, the proposed approach improved the accuracy of the estimates of PTM fold changes and produced a better calibrated false positive rate of detecting differentially abundant PTMs \todo{as compared to methods XXX}. In particular, accounting for the confounding from protein abundance allowed us to characterize the true effect of the modification, avoiding the need for more manual and time intensive follow-up investigation.

The proposed approach is implemented as a freely available open source R package $MSstatsPTM$, as part of the $MSstats$ family of packages \cite{Choi:2014,Huang:2020}, and is available on Bioconductor.

\section{Experimental procedures}

%%%% Summary of controlled mixtures
\begin{table}[t!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of bio. &  Data \\
&  & design & Conditions & replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - Missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Experiment}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets.} 
```Dataset'' is the dataset code name.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode. \todo{Could you add to the table the number of proteins, and the average number of modified/unmodified peptides per protein} \todo{If the figures are at the end, I think the table should also be at the end}
\todo{Could you check the table and the entire section: when we say 'number of biological replicates', we typically mean per condition. But it seems that some descriptions imply the total number of biological samples across conditions. Could you clarify that we use number of biological replicates per condition throughout? (E.g, this helps calculate the degrees of freedom in balanced designs)}
}
\label{tab:dataDescription}
\end{table}

\subsection*{Data overview and availability}

Table \ref{tab:dataDescription} summarizes the experiments. Two computer simulations had known ground truth, and varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. Details of computer simulations are available on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

One spike-in experiment also had known changes in modified spike-in peptides, but had real-world experimental characteristics. Finally, three biological experiments demonstrated the applicability of the proposed approach across different biological organisms, modifications, experimental designs and acquisition strategies. Details of the experimental data, R scripts with $MSstatsPTM$ analysis, and results of the statistical analysis are available in MassIVE.quant (\url{https://massive.ucsd.edu/ProteoSAFe/static/massive-quant.jsp}) \cite{Choi:2020}.

\todo{Additional details on every dataset are available in Supplementary Section 2 (I think it's ok to only mention this once)}

\todo{Could you delete this paragraph, and move the container identifies to a new column in table 1?} The R scripts and results of the statistical analysis for the four experiments are available in MassIVE.quant reanalysis containers: SpikeIn benchmark - Ubiquitination - Label-free (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{TBD}), Human - Ubiquitination - 1mix-TMT (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=b639fa5d0ff24ce78abdca13f4b18c0a}{RMSV000000356}), Mouse - Phosphorylation - 2mix-TMT time series (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=79406d32c1f9413daf7756e6eeb24968}{RMSV000000357}), Human - Ubiquitination - Label-free no global profiling run (MassIVE reanalysis container identifier: \href{https://massive.ucsd.edu/ProteoSAFe/reanalysis_container.jsp?task=f9e6117e00794ce5b2e8802601be16bc}{RMSV000000358}).

\subsection*{Dataset 1: Computer simulation 1 - Label-free}
\label{sec:comp_sim_procedure1}

{\bf Simulation design:} The simulation represented an idealistic case. 24 synthetic label-free datasets were generated with different experimental designs and different biological variation. In each dataset, 1000 proteins had 10 unmodified features per protein. Each of the 1000 proteins had one PTM. Each PTM was represented by 10 modified features. The PTMs of 500 proteins had a differential fold change between conditions, while the other 500 proteins were generated with no changes in abundance between conditions. Furthermore, the fold changes of half of the 500 differential PTMs were fully masked by changes in the unmodified protein. Additionally, the fold change of half the 500 non-differential PTMs was entirely due to changes in the unmodified protein. All the differential PTMs were generated with an expected log fold change of 0.75 between conditions.

Each simulation was generated with random biological variation. The observed peptide abundances were simulated by adding random noise $\mathcal{N}(0,\sigma^2)$ to the deterministic abundances described above. Two values $\sigma^2 = \{.2, .3\}$ were motivated by the experimental datasets in this manuscript.

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods \todo{XXX - what and how}

\subsection*{Dataset 2: Computer simulation 2 - Label-free missing values and low features}
\label{sec:comp_sim_procedure2}

{\bf Simulation design:} The data were simulated as above, while providing a more realistic representation of the experiments. The feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. Specifically, PTMs were simulated with 2 modified peptide features, and unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random.

\medskip \noindent {\bf Evaluation:} \todo{As above, we evaluated... what and how...}

\subsection*{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}

\medskip \noindent {\bf Experimental Design:} Figure \ref{fig:benchmark-design}(a) overviews the experimental design. Four mixtures (i.e., conditions) were created with varying amounts of human lysate, background {\it E. coli} lysate, and human spike-in ub-peptide mixture. Unmodified peptides from human lysate were viewed as the global proteome.   Background {\it E. coli} lysate were used to equalize total protein levels prior to enrichment or global protein profiling \ov{between Mix 1 and 3, and between Mix 2 and 4}.  50 heavy-labeled KGG motif peptides from 20 human proteins were used as spike-in peptides. Quantitative changes in protein and site abundance of these 20 proteins were the target of the benchmark. \todo{In particular, we distinguish the changes in the abundances of the modified peptides (i.e., unadjusted changes) and the changes of their abundance relative to the changes in the abundances of the human lysate (i.e., protein-level adjusted changed). } The true log-fold changes \ov{between the relevant components of the relevant mixtures} are summarized in Figure \ref{fig:benchmark-design}(b).  Two replicates mixtures were created per condition.

\medskip \noindent {\bf Data acquisition:} Each mixture was analyzed with KGG enrichment, and without KGG enrichment (i.e., in a global profiling run), with label-free LC-MS/MS. There was a 90.2\% overlap between the identified background modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Evaluation:} \todo{Here is the place to describe: 'we expect XXX to not change'. Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity). We expect XXX to have fold changes XXX between conditions XXX. We evaluated the accuracy of estimation of fold changes... and the sensitivity of detecting differentially abundant...}  
%
\todo{I think some mixtures are not comparable because the total protein amount is not the same. I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable. I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome. Or was it normalized based on what we know is the same? If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins... estimation... The adjusted proteins... estimation... Positive controls... Negative controls...}

\subsection*{Dataset 4: Human - Ubiquitination - 1mix-TMT}
\label{sec:exp_proc_dataset4}

\medskip \noindent {\bf Experimental Design:}  Luchetti et al. \cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter. Uninfected cells were measured at 0 and 6 hours, while cells infected \todo{with what? Not clear about the infection} were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. \todo{Is GSDMD a known protein that will be used for validation?}

\medskip \noindent {\bf Data acquisition:} 11 samples \todo{why 11 and not 12? Which condition didn't have a biological replicate?} were allocated to 1 TMT mixture in an unbalanced group comparison design. The ubiquitinated peptides, and the total proteome (i.e., global profiling) were each conducted in a single LC-MS/MS run. There was a 95\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

\medskip \noindent {\bf Evaluation:} We evaluated the ability of the statistical methods to detect changes in the abundance of modified peptides both before and after adjusting for changes in global protein abundance. The six condition were labeled Dox1hr, Dox2hr, Dox4hr, Dox6hr, NoDox0hr, and NoDox6hr. All conditions were compared with each other, resulting in 15 pairwise comparisons. Since the dataset was a biological investigation, the true positive modifications were unknown.

\subsection*{Dataset 5: Mouse - Phosphorylation - 2mix-TMT time series}
\label{sec:exp_proc_dataset5}

\medskip \noindent {\bf Experimental Design:} Maculins et al. \cite{Maculins} studied primary murine macrophages infected with {\it Shigella flexneri} ({\it S. flexneri}). The experiment quantified the abundance of total protein and of phosphorylation in wild type (WT), and in ATG16L1-deficient (cKO) samples, uninfected and infected with {\it S. flexneri}. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours).

\medskip \noindent {\bf Data acquisition:} 8 biological samples \todo{total? Or per condition? I am having difficulties understanding the allocation from the description...} were allocated to 2 TMT mixtures in an unbalanced time course design. Each mixture contained two replicates per early and late WT and KO conditions. The six conditions were split between 11 channels, and as the result the experiment was unbalanced. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. This experiment included a total proteome (i.e., a global profiling run) and a phosphopeptide enrichment run. There was a 90\% overlap between the identified modified peptides and proteins that were quantified in the global profiling run.

@@

\subsection*{Dataset 6: Human - Ubiquitination - Label-free no global profiling run}
\label{sec:exp_proc_dataset6}
